---
title: "Week 2 Global Health"
author: "Dr Richard Clarke"
date: "2022-12-18"
output: webexercises::webexercises_default
editor_options: 
  chunk_output_type: console
---

```{r}
library(webexercises)
```

# Week 2

## Exploring Global Life Expectancy

In this weeks R-Lab I want to introduce you to the data visualization package `ggplot`, another package contained within the `tidyverse`. `ggplot` is an incredibly versatile way of creating graphs and has become the standard throughout academic publishing and data journalism (here is a [link](https://medium.com/bbc-visual-and-data-journalism/how-the-bbc-visual-and-data-journalism-team-works-with-graphics-in-r-ed0b35693535) to a post from the BBC talking about how they recently moved to using `ggplot`. Soon you'll be spotting figures everywhere and saying "I bet I could make that with ggplot", and then get sad when you find out they didn't share the data :-(   

This week you will need to following package: 

```{r, echo=TRUE, message=FALSE, error=FALSE, warning=FALSE}

library(tidyverse) # contains dplyr (see last week) and ggplot
library(ggrepel) # used for adding labels to data points
```

Remember if you are using any of these packages for the first time you will need to run `install.packages("packagename")`. It is often best to run this in the console just so you're not reinstalling the package each time you run your code. 

## Todays data

This week we'll be using the global life expectancy dataset that we wrangled last week. 

If you saved your final file last week try to import it using the `read_csv` function. If you didn't save it, below you'll see the code needed to import a cleaned version of the file from my github.  

```{r, echo=TRUE, message=FALSE, error=FALSE, warning=FALSE}
data <- read_csv("https://raw.githubusercontent.com/RichClarkePsy/Datasets/main/Life%20expectancy%20dataset.csv") 
```

Reminder: This dataset is a country level summary of World Health Organisation data on health factors and life expectancy ([source](https://www.kaggle.com/datasets/kumarajarshi/life-expectancy-who)). Data is collected each year but for the purposes of this task we have just selected the data from 2015.

## Visual exploration of the data

Statistics are great, but often, when we want to get a quick and easy understanding of a dataset, we can get everything we want from simple data visualisation. 

Histograms are a good first step to understanding a continuous variable like life expectancy. To create a histogram this we're going to take our dataset then we're going to "pipe" `%>%` it in to the `ggplot` function. 

Each ggplot visulisation is made up of layers. You start by mapping an aesthetic (`aes`), which involves the organisation of what data goes on what axis and how data is segmented within the visualisation. Then you layer on (using a + sign) a geometry to combine the data with a desired visual format. The following example places life expectancy on the x-axis and uses the histogram geometry. 

```{r, echo=TRUE, message=FALSE, error=FALSE, warning=FALSE}
data %>%
  ggplot(aes(x=Life_exp)) +
  geom_histogram()
```

This is fine for our quick exploration of the data but we could also tidy this up a little and it would be perfectly fine to include it in a paper. 

Run the code below and then adapt it to work out what each of the arguments control. Colours can be in words or in hex codes. Here's a useful website where you can get hex codes [www.htmlcolorcodes.com](https://htmlcolorcodes.com/).  

```{r, echo=TRUE, message=FALSE, error=FALSE, warning=FALSE}
data %>%
  ggplot(aes(x=Life_exp)) +
  geom_histogram(binwidth = 1, 
                 colour = "black", 
                 fill = "#58A139") +
  labs(title = "Figure 1. Histogram of Global Life Expectancy",
       subtitle = "N = 183 countries",
       x = "Life expectancy from birth",
       y = "Number of countries",
       caption = "Year: 2015 | Data source: WHO") +
  theme_light() +
  theme(plot.caption = element_text(hjust = 0.5))
  
```

`hide(None of this is magic btw)`
At some point or other I've Googled how to do pretty much every element here.

Try searching the following questions and see if you can find similar looking code:

* "Change the colour of histogram bars in ggplot"
* "Add a caption to a ggplot figure"
* "Align caption to center in ggplot"

The last one here is something I searched for just now while writing this exercise. And no doubt I'll forgetting it and end up having to Google it again in the future.
`unhide()`

That's a good start, but say we want to understand the distribution for each region. We could `filter` the data by each region and create six different histograms. Or we could add the last line of the code below a create a nicely formated version of the six histograms together. 

```{r, message=FALSE, error=FALSE, warning=FALSE}
data %>%
  ggplot(aes(x=Life_exp)) +
  geom_histogram(binwidth = 1, 
                 colour = "black", 
                 fill = "#58A139") +
  labs(title = "Figure 1. Histogram of Global Life Expectancy",
       subtitle = "N = 183 countries",
       x = "Life expectancy from birth",
       y = "Number of countries",
       caption = "Year: 2015 | Data source: WHO") +
  theme_light() +
  theme(plot.caption = element_text(hjust = 0.5)) +
  facet_wrap(~ Continent)
```

`hide(Note about ~)`
The `~` is often used in R to denote a relationship. You'll often see it in regression analysis but in this case it's telling `ggplot` to facet (i.e. seperate) the graph by Continent 
`unhide()`

This visualisation definitely has its value (for instance we can now see the difference in life expectancy between Europe and Africa very clearly), but a better way to visualise the same data might be to change from a histogram to a boxplot.

```{r, message=FALSE, error=FALSE, warning=FALSE}
data %>%
  ggplot(aes(x=Life_exp, y=Continent)) +
  geom_boxplot() +
  theme_light()
```

Much better, instantly we can see how the distribution of life expectancy differs across continents. A few things spring to mind to make this figure more informative. 

* Label the outliers (it'd be nice to know what countries they are)
* Add some colour
* Add title etc



```{r, message=FALSE, error=FALSE, warning=FALSE}
findoutlier <- function(x) {
  return(x < quantile(x, .25) - 1.5*IQR(x) | x > quantile(x, .75) + 1.5*IQR(x))
} # Dont worry too much about this. Basically, what this does is creates a new function that can be used to identify outliers in a variable.   

data_LE <- data %>% 
  group_by(Continent) %>%
  mutate(outlier_LE = ifelse(findoutlier(Life_exp), Country, NA)) # This uses the above function to make a new variables where ever time it finds an outlier in Life_exp it assigns the text from the Country variable. 

# Also becuse I've done some sort of mutation I've changed the name of dataframe. this is so I always have the raw data in the same form as it was read in. Just in case I need to use it again later.  

data_LE %>%
  ggplot(aes(x=Life_exp, y=Continent, fill = Continent)) +
  geom_boxplot(width = 0.5,
               alpha = 0.75,
               show.legend = FALSE) +
  geom_text_repel(aes(label=outlier_LE),
                  position = "identity",
                  size = 3,
                  na.rm=TRUE) +
  scale_fill_brewer(palette = "Dark2") +
  labs(title = "Figure 2. Histogram of Global Life Expectancy",
       subtitle = "N = 183 countries",
       x = "Life expectancy from birth",
       y = "",
       caption = "Year: 2015 | Data source: WHO") +
  theme_light() +
  theme(plot.caption = element_text(hjust = 0))
```

Now that is a good looking figure!

### Exercise

Adapt the code above to make the histograms and boxplots again, but this time for the `BMI` variable instead of life expectancy. 

Note: Here is the code for calculating the outliers for BMI, as this requires adding in the extra line `drop_na(BMI)`. This tells R to remove any missing data in the variable `BMI`. Note the change in dataset name as well.    

```{r, include=FALSE}
findoutlier <- function(x) {
  return(x < quantile(x, .25) - 1.5*IQR(x) | x > quantile(x, .75) + 1.5*IQR(x))
}

data_BMI <- data %>%
  drop_na(BMI) %>%
  group_by(Continent) %>%
  mutate(outlier_BMI = ifelse(findoutlier(BMI), Country, NA)) 
```

What do you notice about the data from this visual check? 

hide(Hint)
I'm concerned about those low BMI data points, BMI should never really be that low. Perhaps this is a data recording issue. No way to know without the raw data to go into. For further analysis we may want to filter out these data points, or at least check them against other data sources. 
unhide()


## Correlation between variables

Next lets take a look at how good `ggplot` is for making scatter plots. This time we're going to plot number of years of schooling (on the x-axis) against life expectancy (on the y=axis) for each of our 183 countries. 

```{r, message=FALSE, error=FALSE, warning=FALSE}
data %>%
  ggplot(aes(x = Schooling, y = Life_exp)) +
  geom_point()
``` 

A very clear correlation. Next week we'll look at how strong this correlation is, but for now we can add the regression line with the geom_smooth function.  

```{r, message=FALSE, error=FALSE, warning=FALSE}
data %>%
  ggplot(aes(x = Schooling, y = Life_exp)) +
  geom_point() +
  geom_smooth(method = "lm",
              formula = y ~ x) +
  labs(title = "Figure 2. A scatter plot of life expectancy by number of years of schooling",) +
  xlab(label = "Average number of years of schooling")+
  ylab(label = "Average life expectancy (in years)")
```

What happens when you change the order and put the geom_smooth before the geom_point function?

There are usually multiple differnt ways to acheive what you with using ggplot. 

Can you delete the bottom two lines of code and find a way to label the x and y axis within the labs function?

hide(Hint)
Check the code from earlier
unhide()



```{r, message=FALSE, error=FALSE, warning=FALSE}
data %>%
  ggplot(aes(x = Schooling, y = Life_exp, colour = Continent)) +
  geom_point()
``` 

Scatter plot seperated into continents by colour and size of the point represents the countrys population.

```{r, message=FALSE, error=FALSE, warning=FALSE}
data %>%
  ggplot(aes(x = Schooling, y = Life_exp, colour = Continent, size = Population)) +
  geom_point()
``` 


Add a linear regression for each continent

```{r}
data %>%
  ggplot(aes(x = Schooling, y = Life_exp, colour = Continent)) +
  geom_point() +
  geom_smooth(method = "lm",
              formula = y ~ x) +
  labs(title = "Figure 2. A scatter plot of life expectancy by number of years of schooling") +
  xlab(label = "Average number of years of schooling")+
  ylab(label = "Average life expectancy (in years)")
``` 

we went too far with this. facet_wrap might be the better way to go. 

```{r}

data %>%
  ggplot(aes(x = Schooling, y = Life_exp)) +
  geom_point() +
    geom_smooth(method = "lm",
              formula = y ~ x) +
  labs(title = "Figure 2. A scatter plot of life expectancy by number of years of schooling") +
  xlab(label = "Average number of years of schooling")+
  ylab(label = "Average life expectancy (in years)") +
  facet_wrap(~ Continent)

``` 

## Exercises

Now do the same but with BMI


